1️⃣ Probability at a specific trial (PMF)

Example: A die is rolled 8 times. Find the probability that exactly 2 successes occur (with success probability = 0.4).
```{r}
# Probability of exactly x successes (PMF)
x <- 2
n <- 8
p <- 0.4

y <- dbinom(x, n, p)
y

```

2️⃣ Cumulative probability (≤ a given value)

Example: Probability of 5 or fewer successes in 100 trials with p = 0.4.
```{r}
# Cumulative probability P(X ≤ 5)
y <- pbinom(5, size = 100, prob = 0.4)
y

```
Quantile (inverse binomial distribution)
```{r}
# Quantile function
y <- qbinom(0.4, size = 5, prob = 0.7)
y

```
4️⃣ Generate random samples from a binomial distribution

Example 1:
Generate 100 observations from 
𝐵(15,0.4)
```{r}
# Random sample from B(15, 0.4)
B <- rbinom(n = 100, size = 15, prob = 0.4)
B

```
Take 90 values from the generated sample y without replacement.
```{r}
B <- sample(y, size = 90, replace = FALSE)

```
6️⃣ Plot frequency distribution of the sample
```{r}
# Frequency plot
plot(table(B),
     main = "Frequency Distribution of Sample B",
     xlab = "Number of Successes",
     ylab = "Frequency")

```





