---
title: "Final Project Statistical Computing"
author: "Group 5"
date: "2026-01-16"
output: html_document
---

GROUP MEMBERS
-------------
1. MASHAMI Pacifique 101406
2. John Seemore 
3. HITAYEZU James 101381
4. Paulin Nkusi 101397
5. IRADUKUNDA Suavis 101391
6. ⁠Mylene MANIRAKIZA 101403

libraries.

```{r}
library(MASS)
```


RE = Var(θ̂₂) / Var(θ̂₁)
If RE > 1, then θ̂₁ is more efficient.

```{r, error=FALSE}

marks <- c(
6.89,10.35,14.5,15.35,13.4,12.35,13.6,11.3,4.6,15.85,10.15,8.15,8.25,
17.05,15.2,16.2,15.05,16.25,18.5,16.8,15.25,12.75,12.45,10.2,13.3,8.7,
13.05,10.9,13,13.85,11.55,14.5,12.85,17.15,13.3,15.25,17.55,11.25,12.85,
14.5,10.7,11.9,14,14.6,12.4,15.35,12.3,11.7,11.7,10,11.8,15.6,
8.6,10,12.5,9.9,10.3,13.3,14,16.5,14,14.7,18.2,15.1,14.4,
8.1,12.4,10.9,13.1,14.5,15.4,10,13.2,13.4,14.6,17.5,13.4,15.8,
13.2,13.3,13.8,11.2,15.3,16.1,9.3,16.5,11.5,13,14.7,9.3,8.6,
12.4,12.8,10.8,5.5,13.5,12.4,11.8,12.2,8.5,9.4,11.9,8.6,13.2,
12.4,11.5,10.1,14.8,12.7,6.1,8.6,9.3,9.9,8.4,11.3,13.1,13.5,
4.1,7.5,13.9,15.1,13.2,14.5,17.6,14.7,12.5,15.4,15.5,9.2,10.5,
13.45,18.15,10.9,12.65,13.2,17.6,13.9,12.4,14.6,12.4,17.15,16.4,13.75,
14.4,17.35,14.3,14.65,16.1,11.25,10.35,13.95,12.55,7.6,14.95,9.6,18.35,
12.2,12.35,16.7,15.3,9,14.05,14.65,11.15,17.45,4.75,13.6,14.4,15.2,
8.75,13.4,13.6,14.35,12.65,14.9,15.1,15.45,16.55,9.75,16.9,14.35,12.66,
11.44,11.47,12.1,11.34,10.88,8.34,13.24,15.59,14.09,12.4,15.84,12.39,17.64,
12.78,12.78,15.41,11.21,12.56,10.85,16.04,11.1,13.25,15.9,14.5,12.42,11,
10.08,12.8,8.18,15.75,13.26,12.61,13.98,6.77,13.85,12.42,15.36,13.24,13.88,
17.7,15.02,17.22,16.98,16.12,17.66,17.64,13.32,17.14,11.6,18.1,14.13,14.98,
16.44,18.73,14.82,16.02,14.66,14.48,12.56,10.21,18.72,19.42,16.22,17.07,15.82,
15.27
)

```

Exploratory analysis
X∼N(μ,σ2)

```{r ,error=FALSE}
summary(marks)
hist(marks, breaks = 20, probability = TRUE)
```

Statistical reasoning: 

1.Data are continuous
2.Values cluster around a central mean
3.Histogram is approximately symmetric
4.No strong skewness
5.According to Rizzo, such data are often modeled by a Normal distribution

Conclusion: X∼N (μ,σ2)

Find the analytical form of the distribution
Estimate parameters

```{r, error=FALSE}
mu_hat <- mean(marks)
sigma_hat <- sd(marks)

mu_hat
sigma_hat

```
Analytical density
f(x)=1/2πσ²^1/2 exp(-(x−μ)2/2σ²)

μ=xˉ,σ=s

```{r, error=FALSE}
set.seed(123)

n <- length(marks)
simulated_marks <- rnorm(n, mean = mu_hat, sd = sigma_hat)

```


Generate a random sample of the same size


```{r, error=FALSE}
set.seed(123)

n <- length(marks)
simulated_marks <- rnorm(n, mean = mu_hat, sd = sigma_hat)

```


Goodness-of-fit test
We use Kolmogorov–Smirnov test

```{r, error=FALSE}
ks.test(marks, "pnorm", mean = mu_hat, sd = sigma_hat)

```
Interpretation

H₀: Data follow Normal(μ, σ²)

If p-value > 0.05 → do not reject H₀

QUESTION 2

Multivariate normal simulation from rock dataset

μ=E(X)

```{r, error=FALSE}
data(rock)
head(rock)

mu_hat <- colMeans(rock)
mu_hat

```
Σ=Cov(X)
```{r}
Sigma_hat <- cov(rock)
Sigma_hat
```
We assume That X∼Np​(μ,Σ)

where:
p = number of variables
μ=μ^
Σ=Σ^

```{r, error=FALSE}

set.seed(123)

n <- 500

simulated_rock <- mvrnorm(
  n = n,
  mu = mu_hat,
  Sigma = Sigma_hat
)

```

Convert to data frame and inspect
```{r}
simulated_rock <- as.data.frame(simulated_rock)
head(simulated_rock)

```
The last six records

```{r}
tail(simulated_rock)
```



QUESTION 3

The Rayleigh distribution with parameter σ>0 has density:
f(x)=x/σ². e^−x^2/(2σ²),x≥0

Find the cumulative distribution function
Integrate the density:
F(x)=∫0<->x t/σ²e−t^2/(2σ²) dt

F(x)=1−e^−x2/(2σ²)


Apply inverse transform method:
Let U∼Uniform(0,1)

Set: U=1−e^−x2/(2σ²)

Solve for X:

X= σ(-2ln(1−U)^1/2


Algorithm
1.Generate U∼U(0,1)

2.Compute X= σ(-2ln(1−U)^1/2


```{r, error=FALSE}
set.seed(123)

n <- 1000
sigma <- 2

u <- runif(n)
x <- sigma * sqrt(-2 * log(1 - u))

hist(x, probability = TRUE, breaks = 30,
     main = "Rayleigh Distribution")

curve((x / sigma^2) * exp(-x^2 / (2 * sigma^2)),
      from = 0,
      to = max(x),
      add = TRUE,
      col = "red",
      lwd = 2)

```


Accept–Reject Method

Target density
Consider a triangular distribution on [0,1] (Triangular Distribution)

f(x)= {2x,0≤x≤1 ,  0,otherwise 

Proposal distribution:
Choose:g(x)=1,0≤x≤1   which is U(0,1)
We need M such that: f(x)≤Mg(x)

Since 
max f(x)=2, choose M=2

Accept–Reject rule
1.Generate X∼U(0,1)
2. Generate U∼U(0,1)
3. Accept X if: U<= f(x)/Mg(x)=2x/2= x

```{r, error=FALSE}
set.seed(123)

n <- 1000
res <- numeric(n)
i <- 1

while (i <= n) {
  x <- runif(1)
  u <- runif(1)
  if (u <= x) {
    res[i] <- x
    i <- i + 1
  }
}

hist(res, probability = TRUE, breaks = 30,
     main = "Triangular Distribution")

curve(2 * x,
      from = 0,
      to = 1,
      add = TRUE,
      col = "red",
      lwd = 2)

```


Sum and Mixture Method

(Mixture of Two Normal Distributions)

Define the mixture model
let x={N(-2,1), with probability 0.4,   N(2,1), with probability 0.6


Algorithm
1. Generate B∼Bernoulli(0.4)
2. If B=1, generate from N(−2,1)
3. Otherwise, generate from N(2,1)

```{r}
set.seed(123)

n <- 1000
b <- rbinom(n, 1, 0.4)

x <- ifelse(b == 1,
            rnorm(n, -2, 1),
            rnorm(n, 2, 1))

hist(x, probability = TRUE, breaks = 40,
     main = "Mixture of Normals")


curve(0.4 * dnorm(x, -2, 1) + 0.6 * dnorm(x, 2, 1),
      from = min(x),
      to = max(x),
      add = TRUE,
      col = "red",
      lwd = 2)
```

